---
title: "Spatial Voting Analysis with socialchoicelab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Voting Analysis with socialchoicelab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # requires libscs_api; set eval=TRUE with SCS_LIB_PATH set
)
```

> **Note:** All code chunks require `libscs_api` to be compiled and
> `SCS_LIB_PATH` to point to the build directory. See the package
> [README](https://github.com/SocialChoiceStudio/socialchoicelab) for build
> instructions.

```{r load}
library(socialchoicelab)
```

## 1 — Voter ideal points in 2D

We model a legislature with **5 voters** whose ideal policies live in a
two-dimensional issue space (e.g., economic left–right × social liberal–conservative).

```{r voters}
# Flat [x0, y0, x1, y1, ...] representation — 5 voters.
voters <- c(
  -1.0, -0.5,   #  voter 1: left-economic, moderate-social
   0.0,  0.0,   #  voter 2: centrist
   0.8,  0.6,   #  voter 3: right-economic, conservative
  -0.4,  0.8,   #  voter 4: centre-left, liberal
   0.5, -0.7    #  voter 5: right-economic, libertarian
)

# Three policy alternatives: status quo SQ, reform A, reform B.
sq <- c( 0.0,  0.0)   # status quo at centroid
a  <- c( 0.6,  0.4)   # right-of-centre reform
b  <- c(-0.5,  0.3)   # left-of-centre reform
```

## 2 — Reproducible randomness with StreamManager

`StreamManager` encapsulates a master seed and named sub-streams so that all
stochastic draws are fully reproducible.

```{r stream-manager}
sm <- stream_manager(master_seed = 2024, stream_names = c("gen", "ties"))
# Draw a test value — for illustration only.
sm$uniform_real("gen", 0, 1)
```

## 3 — Majority preference between two alternatives

Does a simple majority of voters prefer reform A to the status quo?

```{r majority}
majority_prefers_2d(a, sq, voters)   # TRUE/FALSE
majority_prefers_2d(b, sq, voters)
```

## 4 — Winset: which policies beat the status quo?

The **winset** of the status quo is the set of all policy points that a strict
majority of voters prefer to `sq`. An empty winset means the status quo is a
majority-rule equilibrium.

```{r winset}
ws <- winset_2d(status_quo = sq, voter_ideals = voters)
ws$is_empty()        # TRUE if SQ is a majority equilibrium
ws$bbox()            # bounding box of the winset
```

Plot the winset boundary (requires **graphics**):

```{r winset-plot, fig.width = 5, fig.height = 5}
bnd <- ws$boundary()
plot(bnd$xy, type = "n", asp = 1, xlab = "x", ylab = "y",
     main = "Winset of the status quo")
polygon(bnd$xy[, 1], bnd$xy[, 2], col = "#CCE5FF88", border = "#336699")
points(rbind(sq, a, b), pch = 19, col = c("black", "firebrick", "steelblue"))
text(rbind(sq, a, b), labels = c("SQ", "A", "B"), pos = 4)
points(matrix(voters, ncol = 2, byrow = TRUE), pch = 4, col = "grey40")
legend("topright", legend = c("Winset", "Alternatives", "Voters"),
       fill = c("#CCE5FF88", NA, NA), pch = c(NA, 19, 4),
       col = c(NA, "firebrick", "grey40"), border = c("black", NA, NA))
```

## 5 — Copeland winner

The **Copeland winner** beats (or ties) the most alternatives in pairwise
majority comparisons. With a finite alternative set:

```{r copeland}
alts <- c(sq, a, b)   # flat representation
copeland_scores_2d(alts, voters)   # named score vector
copeland_winner_2d(alts, voters)   # 1-based index of winner
```

## 6 — Condorcet winner

A **Condorcet winner** beats *every* other alternative by strict majority.

```{r condorcet}
has_condorcet_winner_2d(alts, voters)       # logical
condorcet_winner_2d(alts, voters)           # 1-based index or NA
```

## 7 — Ordinal preference profiles and voting rules

Build an ordinal profile from the spatial layout, then apply plurality and
Borda count:

```{r profile}
prof <- profile_build_spatial(alt_xy = alts, voter_ideals = voters)
prof$dims()           # list(n_voters = 5, n_alts = 3)
prof$export()         # 5×3 ranking matrix
```

**Plurality:**

```{r plurality}
plurality_scores(prof)
plurality_one_winner(prof)
```

**Borda count:**

```{r borda}
borda_scores(prof)
borda_ranking(prof)    # full social ordering
```

**Approval voting** (each voter approves their top 2):

```{r approval}
approval_scores_topk(prof, k = 2L)
approval_all_winners_topk(prof, k = 2L)
```

## 8 — Pareto efficiency and Condorcet consistency

```{r properties}
pareto_set(prof)                  # Pareto-efficient alternatives (1-based)
has_condorcet_winner_profile(prof)
condorcet_winner_profile(prof)    # 1-based index or NA

# Check whether alt 1 is Condorcet-consistent:
is_condorcet_consistent(prof, 1L)
```
