---
title: "Spatial Voting Analysis with socialchoicelab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Voting Analysis with socialchoicelab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # requires libscs_api; set eval=TRUE with SCS_LIB_PATH set
)
```

> **Note:** All code chunks require `libscs_api` to be compiled and
> `SCS_LIB_PATH` to point to the build directory. See the package
> [README](https://github.com/SocialChoiceStudio/socialchoicelab) for build
> instructions.

```{r load}
library(socialchoicelab)
```

## 1 — Voter ideal points in 2D

We model a legislature with **5 voters** whose ideal policies live in a
two-dimensional issue space (e.g., economic left–right × social liberal–conservative).

```{r voters}
# Flat [x0, y0, x1, y1, ...] representation — 5 voters.
voters <- c(
  -1.0, -0.5,   #  voter 1: left-economic, moderate-social
   0.0,  0.0,   #  voter 2: centrist
   0.8,  0.6,   #  voter 3: right-economic, conservative
  -0.4,  0.8,   #  voter 4: centre-left, liberal
   0.5, -0.7    #  voter 5: right-economic, libertarian
)

# Three policy alternatives: status quo SQ, reform A, reform B.
sq <- c( 0.0,  0.0)   # status quo at centroid
a  <- c( 0.6,  0.4)   # right-of-centre reform
b  <- c(-0.5,  0.3)   # left-of-centre reform
```

## 2 — Reproducible randomness with StreamManager

`StreamManager` encapsulates a master seed and named sub-streams so that all
stochastic draws are fully reproducible.

```{r stream-manager}
sm <- stream_manager(master_seed = 2024, stream_names = c("gen", "ties"))
sm$uniform_real("gen", 0, 1)
```

## 3 — Majority preference between two alternatives

Does a simple majority of voters prefer reform A to the status quo?

```{r majority}
majority_prefers_2d(a, sq, voters)   # TRUE/FALSE
majority_prefers_2d(b, sq, voters)
```

## 4 — Winset: which policies beat the status quo?

The **winset** of the status quo is the set of all policy points that a strict
majority of voters prefer to `sq`. An empty winset means the status quo is a
majority-rule equilibrium.

```{r winset}
ws <- winset_2d(status_quo = sq, voter_ideals = voters)
ws$is_empty()   # TRUE if SQ is a majority equilibrium
ws$bbox()       # bounding box of the winset
```

## 5 — Copeland and Condorcet winners

The **Copeland winner** beats (or ties) the most alternatives in pairwise
majority comparisons. A **Condorcet winner** beats *every* alternative by
strict majority.

```{r copeland-condorcet}
alts <- c(sq, a, b)   # flat representation
copeland_scores_2d(alts, voters)           # named score vector
copeland_winner_2d(alts, voters)           # 1-based index
has_condorcet_winner_2d(alts, voters)      # logical
condorcet_winner_2d(alts, voters)          # 1-based index or NA
```

## 6 — Centrality measures

Three summary statistics for the voter ideal-point distribution:

```{r centrality}
marginal_median_2d(voters)    # coordinate-wise median (issue-by-issue median voter)
centroid_2d(voters)           # coordinate-wise arithmetic mean (centre of mass)
# geometric_mean_2d() requires strictly positive coordinates
```

## 7 — Ordinal preference profiles and voting rules

Build an ordinal profile from the spatial layout, then apply voting rules:

```{r profile}
prof <- profile_build_spatial(alt_xy = alts, voter_ideals = voters)
prof$dims()     # list(n_voters = 5, n_alts = 3)
prof$export()   # 5×3 ranking matrix (1 = most preferred)
```

**Plurality, Borda count, and approval voting:**

```{r voting-rules}
plurality_scores(prof)
borda_scores(prof)
borda_ranking(prof)                    # full social ordering
approval_scores_topk(prof, k = 2L)    # each voter approves top 2
```

## 8 — Pareto efficiency and Condorcet consistency

```{r properties}
pareto_set(prof)                  # Pareto-efficient alternatives (1-based)
has_condorcet_winner_profile(prof)
condorcet_winner_profile(prof)    # 1-based index, or NULL
is_condorcet_consistent(prof, 1L) # is alternative 1 Condorcet-consistent?
```

## 9 — Built-in scenarios

`socialchoicelab` ships 33 canonical spatial voting scenarios:

```{r scenarios}
list_scenarios()                          # character vector of names
s <- load_scenario("laing_olmsted_bear")  # returns list(voters, status_quo, ...)
s$voters                                  # flat [x0, y0, ...] vector
s$status_quo                              # length-2 vector
```

## 10 — Interactive visualization

All plotting functions return composable **Plotly** figures.

```{r plot-setup}
voter_names <- paste0("P", 1:5)
alts_flat   <- c(sq, a, b)

fig <- plot_spatial_voting(
  voters,
  alts_flat,
  sq          = sq,
  voter_names = voter_names,
  alt_names   = c("SQ", "A", "B"),
  title       = "5-voter legislature"
)
fig
```

Add geometric layers one at a time — each returns the same figure enriched
with a new trace:

```{r plot-layers}
hull <- convex_hull_2d(voters)
ws   <- winset_2d(sq, voters)
bnd  <- uncovered_set_boundary_2d(voters, grid_resolution = 12L)

fig <- plot_spatial_voting(voters, alts_flat, sq = sq,
                           voter_names = voter_names,
                           alt_names   = c("SQ", "A", "B"),
                           title       = "Spatial analysis — composite layers")
fig <- layer_convex_hull(fig, hull)
fig <- layer_winset(fig, ws)
fig <- layer_uncovered_set(fig, bnd)
fig <- layer_ic(fig, voters, sq, color_by_voter = TRUE)
fig <- layer_centroid(fig, voters)
fig <- layer_marginal_median(fig, voters)
fig <- finalize_plot(fig)
fig
```

### Color themes

All `layer_*()` functions accept a `theme` argument for colorblind-safe palettes:

```{r themes}
scl_palette("okabe_ito", n = 5)  # return hex colors from the Okabe-Ito palette
scl_theme_colors("winset")       # canonical color for the winset layer
```

### Saving plots

```{r save}
save_plot(fig, "analysis.html")           # self-contained HTML
# save_plot(fig, "analysis.png")          # raster image (requires kaleido)
```
